---
title: "Proba Sim 24 - Session 9"
author: "Davide Legacci"
date: "`r Sys.Date()`"
output: html_document
---

- c.f. https://rpubs.com/janselmi/TD2 for exercises and code of this session
- c.f. https://github.com/jonatha-anselmi/INFO4-PS for course page
- c.f. https://github.com/davidelegacci/probasim24 for tutorial page

## Mr. and Mrs. Smith


### Directly

- Label the possible configurations as $1 = FF, 2 = FM, 3 = MF, 4 = MM$

- Draw $N$ random integers between 1 and 4 with uniform probability, allowing repetition (`replace = TRUE`) 

- every drawn integer is a family configuration (two children, either male or female)


```{r}
N=10

S=sample(x=1:4,size=N,replace=T)
print(S)
```

We want to se how many times

- given that one is a girl (that is, eliminating all the 4s from the list)
- the other is a boy (that is, how many 2s and 3s are left)

```{r}
print(S[S<4])
```

- `S[S < 4]`: This filters the elements in `S` where the condition `S < 4` is `TRUE.` In other words, it keeps only those elements in `S` that are less than 4, that in this example are `r length(S[S<4])` over the initial `r N`.
- Once there are only 1s, 2s and 3s in `S`, we need to calculate the ratios of 2s and 3s.

```{r}
print(S[S<4]!=1)

print(  mean(S[S<4]!=1)  )
```


- `(S[S < 4] != 1` is a logical vector; `TRUE` is evaluated as `1` and `FALSE` as `0`
- `mean` on this logical vector returns the ratios of 2s and 3s in the list, that is `r length(S[S<4 & S > 1])` over `r length(S[S<4])`, i.e. `r length(S[S<4 & S > 1]) / length(S[S<4])`.


Putting all together and running the experiment many  times, 
```{r}
N=100000
S=sample(x=1:4,size=N,replace=T)
```

```{r}
prob=mean(S[S<4]!=1)
print(prob)
```


Alternatively, mirroring more explicitly the formula for the conditional probability $$P(A | B) = \frac{P(A \text{ and } B)}{P(B)}$$
```{r}
prob = length( S[S<4 & S > 1] ) / length( S[S<4] )
print(prob)
```

### Some practice with dataframes



```{r}
# Number of experiments, that is family instances
N = 10
```


- Create a Data Frame where each family (row) is assigned a possible configuration (the column values)
```{r}
df = data.frame(
    C1 = sample(x = c("F", "G"), size = N, replace = TRUE),
    C2 = sample(x = c("F", "G"), size = N, replace = TRUE)
)
```

- This creates a data frame `df` with two columns, `C1` and `C2`, each containing `N` random samples of either `"F"` (female) or `"G"` (male).
- `sample(x = c("F", "G"), size = N, replace = TRUE)` randomly samples `"F"` or `"G"` `N` times, allowing repetitions (`replace = TRUE`), meaning each child can independently be male or female.
- Each row in `df` represents a "family" with two children, where `C1` is the gender of the first child and `C2` is the gender of the second child.

```{r}
print(df)
```


- next, filter the families with at Least one female
```{r}
df = df[df$C1 == "F" | df$C2 == "F", ]
```

- This line filters `df` to include only families where **at least one child is female**.
- `df$C1 == "F" | df$C2 == "F"` creates a logical vector that is `TRUE` for rows where `C1` **or** `C2` is `"F"`.
- `df[...]` subsets `df` to include only the rows that meet this condition.
- After this line, `df` contains only families with at least one female child.

```{r}
print(df)
```

- next calculate the probability of having a boy in these families
```{r}
prob = nrow(df[df$C1 == "G" | df$C2 == "G", ]) / nrow(df)
```

- `df[df$C1 == "G" | df$C2 == "G", ]` filters `df` again, but this time it keeps only rows where at least one child is male (`C1 == "G" | C2 == "G"`).
- `nrow(df[df$C1 == "G" | df$C2 == "G", ])` gives the number of families with at least one boy, among those with at least one female
- `nrow(df)` gives the total number of families with at least one female
- The division `nrow(...) / nrow(...)` calculates the probability that a family  with at least one female also has a male, that is

```{r}
prob
```

Putting all together and letting `N` big,
```{r}
N=1000000
df=data.frame(C1=sample(x=c("F","G"),size=N,replace=T),
              C2=sample(x=c("F","G"),size=N,replace=T));
df=df[df$C1=="F" | df$C2=="F",] # only families with at least a female
prob=nrow(df[df$C1=="G" | df$C2=="G",])/nrow(df) # how many families with a boy (among the ones with at least a female)
prob
```

---

## A winner among K
$\newcommand{\winner}{i^{\ast}}$
$\newcommand{\players}{K}$
Along the lines of the solution in https://rpubs.com/janselmi/TD2, the sample space is $\Omega = \{0, 1\}^k$ endowed with the uniform probability (with $k \geq 3$). That is, let $\omega = (\omega_i)_{i = 1, \dots, k} \in \Omega$ be the profile of coin tosses, where $\omega_i \in {0, 1}$ is the outcome of player's $i$ toss, for all $i \in \players := \{1, \dots, k\}$. We have $|{\Omega}| = 2^k$, and $p(\omega) = \frac{1}{2^k}$.
The event "there is a winner" is
$$W \subset \Omega = \{\omega \in \Omega: \text{there exists precisely one } \winner \in \players \text{ such that }
\omega_\winner \neq \omega_i \forall i \in \players - \{\winner\} \}
$$
We can enumerate the points in $W$:

- 1: the first player tosses $0$ and all other toss $1$
- 2: the second player tosses $0$ and all other toss $1$
- ...
- k: the $k$-th player tosses $0$ and all other toss $1$
- k+1: the first player tosses $1$ and all other toss $0$
- k+2: the second player tosses $1$ and all other toss $0$
- ...
- 2k: the $k$-th player tosses $1$ and all other toss $0$

Thus we have $|W| = 2k$, and $$p(W) = \frac{2k}{2^k} = \frac{k}{2^{k-1}}$$

**Let $T_k$ be the RV** denoting the number of games to play for observing a winner. What are the law and expectation of this RV?

$\newcommand{pber}{p_k}$
Each play is a Bernoulli event with probability $\pber$, thus by definition $T_k$ follows a geometric distribution with parameter $\pber)$, i.e. for all $t = 1, 2, \dots$ we have

$$
P(T_k = t) = \pber(1-\pber)^{t-1}
$$

with expected value $\frac{1}{\pber}$.

### R code

#### Direct solution


Set the Number of Simulations and Players

```{r}
N = 10
K = 5
```

- for each of the `N` trials generate a game instance, that is a binary vector of size `K`
- represent it as a matrix with `N` rows and `K` columns.
```{r}
S=replicate(K, sample(x=0:1,N,replace=T)) 
print(S)
```


Next, calculate the probability of having exactly `1` or `(K - 1)` ones in each row:
```{r}
p_K = mean(rowSums(S)==1 | rowSums(S)==K-1)
```

- `rowSums(S)` calculates the sum of each row in the matrix `S`, which gives the count of 1s in each row.
- `rowSums(S) == 1 | rowSums(S) == K - 1` is a logical vector that is TRUE for rows where the sum is exactly 1 or K - 1 (in this case, `r sum(rowSums(S) == 1 | rowSums(S) == K - 1)` out of `r N`).
- `mean(...)` calculates the average of this logical vector, effectively giving the probability that a row has exactly 1 or K - 1 ones.

```{r}
p_K
```
Putting all together for `N` large, 
```{r}
N=1000000
K=5
S=replicate(K, sample(x=0:1,N,replace=T)) 
p_K=mean(rowSums(S)==1 | rowSums(S)==K-1)
p_K
```
Notice that the theoretical value is $\frac{`r K` }{ 2^{`r K-1`} } = `r K / (2^(K-1))`$.

---

#### Solution with dataframes

Set the Number of Simulations `N` and Players `K`

```{r}
N = 10
K = 5
```

First, we represent the history of plays of one player as a data frame `df_parties` with one column, called `x1`, and `N` rows, containing random binary values (`0` or `1`).

```{r}
# generates a sample of `0`s and `1`s of length `N`.
df_parties = data.frame(x1 = sample(x = c(0,1), size = N, replace = TRUE))
print(df_parties)
```

Then, we add the history of plays of the other players (that is, we add `K-1` columns of the dataframe):

```{r}
for(i in 2:K) { 
    df_parties = cbind(df_parties, data.frame(x = sample(x = c(0,1), size = N, replace = TRUE)))
    names(df_parties)[i] = paste0("x", i)
}
print(df_parties)
```
- This loop adds `K-1` additional columns to `df_parties`, each filled with random binary values (`0` or `1`).
- In each iteration, we add a new column to `df_parties` using `cbind()` and rename it as `"x2"`, `"x3"`, etc., using `paste0("x", i)`.


**Initialize a Sum Column**

We add a column called `sum` to `df_parties` and initialize it to `0`. This column will store the row-wise sum of the values in each `x` column.
```{r}
df_parties$sum = 0
print(df_parties)
```


**Calculate the Sum of Each Row and store it in the sum column**

```{r}
for(i in 1:K) { 
    df_parties$sum = df_parties$sum + df_parties[, paste0("x", i)]
}
print(df_parties)
```

- This loop iterates over each column from `x1` to `xK`, adding each value to the `sum` column.
- The result is that `df_parties$sum` contains the total number of `1`s in each row.


**Calculate the Probability**

- Finally, calculate the probability that each row has exactly `1` or `K-1` (in this case `r K-1`) ones
- `mean(df_parties$sum == 1 | df_parties$sum == K - 1)` calculates the fraction of rows where `sum` is `1` or `K-1`.

```{r}
p_K = mean(df_parties$sum == 1 | df_parties$sum == K - 1)
p_K
```

Putting all together for `N` large, 
```{r}
N=1000000
K=5
df_parties = data.frame(x1 = sample(x=c(0,1),size=N,replace=T));
for(i in 2:K) { 
    df_parties = cbind(df_parties, data.frame(x = sample(x=c(0,1),size=N,replace=T)));
    names(df_parties)[i]=paste0("x",i);
}
df_parties$sum = 0
for(i in 1:K) { 
    df_parties$sum = df_parties$sum + df_parties[,paste0("x",i)];
}
p_K=mean(df_parties$sum==1 | df_parties$sum==K-1);
p_K
```
Notice that the theoretical value is $\frac{`r K` }{ 2^{`r K-1`} } = `r K / (2^(K-1))`$.
